---
title: "Bayesian Causal Inference via G-Computation: Intuition, Predictive Distributions, and Examples"
author: " "
date: today
format: 
  html:
    theme: cosmo
    css: "/Users/nam-anhtran/heartofsaigon/styles/styles_doc.css"
    toc: true
    code-fold: show
    self-contained-math: true
    embed-resources: true
    html-math-method: katex
    number-sections: true
    include-in-header: "/Users/nam-anhtran/heartofsaigon/styles/noindex.html"
execute: 
  echo: true
  eval: true
  fig-height: 7
bibliography: references.bib
csl: https://www.zotero.org/styles/sage-vancouver-brackets
editor_options: 
  chunk_output_type: console
---

## Overview

These notes explain Bayesian causal inference using **g-computation**. The goal is to clarify the relationship between:

- causal identification via the **g-formula**
- Bayesian inference via the **posterior distribution**
- prediction via the **posterior predictive distribution**

We focus on the practical workflow for analyzing observational data with a binary treatment.

The key idea is simple:

> Fit a Bayesian outcome model, simulate hypothetical outcomes under different treatments, and average over the target population.

This procedure is known as **Bayesian g-computation**.

## The causal question

Suppose we observe data

$$
(Y_i, T_i, X_i), \quad i = 1,\dots,n
$$

where

- $Y$ is the outcome
- $T \in \{0,1\}$ is the treatment
- $X$ represents baseline covariates

The causal question is:

> What would be the average outcome if everyone received treatment 1?  
> What would be the average outcome if everyone received treatment 0?

These quantities are

$$
E(Y \mid do(T=1))
$$

and

$$
E(Y \mid do(T=0)).
$$

The **average treatment effect** is

$$
ATE = E(Y \mid do(T=1)) - E(Y \mid do(T=0)).
$$

## Identification via the g-formula

If treatment assignment is ignorable given $X$, meaning that after conditioning on $X$, the observed treatment assignment carries no additional information about the outcome beyond the treatment received, then the intervention mean is

$$
E(Y \mid do(T=t))
=
\int E(Y \mid T=t, X=x)\, f_X(x)\, dx.
$$

This is the **g-formula**.

Interpretation:

1. compute the mean outcome for each covariate value $x$
2. set treatment equal to $t$
3. average those conditional means over the target population distribution of $X$

If $X$ is discrete, the integral becomes a sum:

$$
E(Y \mid do(T=t))
=
\sum_x E(Y \mid T=t, X=x)P(X=x).
$$

If $X$ is continuous, we use the density $f_X(x)$ or, in practice, the empirical distribution of the observed $X_i$ values.

## Why the g-formula matters

A naive comparison

$$
E(Y \mid T=1) - E(Y \mid T=0)
$$

compares two observed groups that may have different covariate distributions.

The g-formula fixes this by asking:

> What would happen if the **same target population** received treatment 1?  
> What would happen if that **same target population** received treatment 0?

That is why the g-formula is the formal causal ingredient behind the intuitive duplicated-dataset algorithm.

## The Bayesian outcome model

To estimate the conditional outcome law, we specify a Bayesian model such as

$$
Y_i \mid T_i, X_i, \theta \sim p(y \mid T_i, X_i, \theta)
$$

with prior

$$
\theta \sim p(\theta).
$$

After observing data $D$, we obtain the posterior distribution

$$
p(\theta \mid D).
$$

This posterior quantifies uncertainty about the unknown parameters.

## Posterior distribution versus posterior predictive distribution

### Posterior distribution

The posterior distribution is

$$
p(\theta \mid D).
$$

It describes uncertainty about unknown model parameters after seeing the data.

Examples of parameters include:

- regression coefficients
- variances
- subgroup means
- interaction terms

### Posterior predictive distribution

The posterior predictive distribution for a future outcome is

$$
p(y^{new} \mid T=t, X=x, D)
=
\int p(y^{new} \mid T=t, X=x, \theta)p(\theta \mid D)d\theta.
$$

This integrates over posterior uncertainty in $\theta$.

So:

- the **posterior distribution** is about unknown parameters
- the **posterior predictive distribution** is about future outcomes

## How posterior prediction and the g-formula work together

The intervention mean can be written as

$$
E(Y \mid do(T=t), D)
=
\int E(Y \mid T=t, X=x, D)\, f_X(x)\, dx.
$$

Thus:

1. the Bayesian model provides conditional predictions
2. the g-formula tells us how to average those predictions across the target population

The g-formula is therefore **not separate** from the predictive approach. The predictive approach is one practical way to implement the g-formula.

## Intuition: simulate two hypothetical worlds

The algorithm is easy to picture.

### World 1: everyone treated

- keep each subject’s observed covariates $X_i$
- set $T_i = 1$ for all subjects
- predict or simulate outcomes

### World 0: everyone untreated

- keep the same $X_i$
- set $T_i = 0$ for all subjects
- predict or simulate outcomes

Then average the outcomes in each world.

The difference between these two averages is the causal effect.

This is exactly **Bayesian g-computation**.

## Illustrative example with discrete subgroup

Suppose $X$ is a binary subgroup indicator and the target population satisfies

$$
P(X=0)=0.7, \qquad P(X=1)=0.3.
$$

Suppose the fitted model gives

$$
E(Y \mid T=1,X=0)=5,
$$

$$
E(Y \mid T=1,X=1)=8,
$$

$$
E(Y \mid T=0,X=0)=7,
$$

$$
E(Y \mid T=0,X=1)=10.
$$

Then

$$
E(Y \mid do(T=1))
=
0.7(5) + 0.3(8) = 5.9
$$

and

$$
E(Y \mid do(T=0))
=
0.7(7) + 0.3(10) = 7.9.
$$

The treatment effect is

$$
5.9 - 7.9 = -2.
$$

This example shows the subgroup-weighted form of the g-formula.

## Continuous covariate example

Now suppose $X$ is continuous. For example, let $X$ be age and use the outcome model

$$
E(Y \mid T,X)
=
\beta_0 + \beta_1T + \beta_2X + \beta_3TX.
$$

Then for subject $i$,

$$
E(Y \mid T=1,X_i)
=
\beta_0 + \beta_1 + \beta_2X_i + \beta_3X_i,
$$

and

$$
E(Y \mid T=0,X_i)
=
\beta_0 + \beta_2X_i.
$$

The intervention means are approximated by

$$
\psi_1 =
\frac{1}{n}
\sum_{i=1}^n
E(Y \mid T=1,X_i)
$$

and

$$
\psi_0 =
\frac{1}{n}
\sum_{i=1}^n
E(Y \mid T=0,X_i).
$$

This is just the empirical version of integrating over the continuous covariate distribution.

## Example dataset in R

```{r}
set.seed(123)

dat <- data.frame(
  id = 1:6,
  X  = c(20, 35, 50, 40, 60, 30),
  T  = c(1, 0, 1, 0, 1, 0),
  Y  = c(5.1, 7.4, 6.2, 8.0, 7.1, 6.8)
)

dat
```

## Simulated posterior draws

For illustration, we generate artificial posterior draws for a simple linear outcome model

$$
E(Y \mid T,X) = \beta_0 + \beta_1T + \beta_2X.
$$

```{r}
M <- 2000

beta0 <- rnorm(M, 6, 0.4)
beta1 <- rnorm(M, -1.2, 0.3)
beta2 <- rnorm(M, 0.03, 0.01)

n <- nrow(dat)
```

## Bayesian g-computation using conditional means

For each posterior draw, we create two hypothetical worlds:

- all subjects assigned $T=1$
- all subjects assigned $T=0$

and keep each subject’s observed covariate value $X_i$.

```{r}
psi1 <- numeric(M)
psi0 <- numeric(M)

for (m in 1:M) {

  mu1 <- beta0[m] + beta1[m] * 1 + beta2[m] * dat$X
  mu0 <- beta0[m] + beta1[m] * 0 + beta2[m] * dat$X

  psi1[m] <- mean(mu1)
  psi0[m] <- mean(mu0)
}

delta <- psi1 - psi0
```

## Posterior summary of the causal effect

```{r}
c(
  mean_treated = mean(psi1),
  mean_control = mean(psi0),
  mean_effect  = mean(delta),
  lower        = quantile(delta, 0.025),
  upper        = quantile(delta, 0.975)
)
```

## Posterior predictive simulation version

Instead of using conditional means directly, we can simulate outcomes from the posterior predictive distribution.

```{r}
sigma <- abs(rnorm(M, 0.8, 0.1))

psi1_sim <- numeric(M)
psi0_sim <- numeric(M)

for (m in 1:M) {

  mu1 <- beta0[m] + beta1[m] * 1 + beta2[m] * dat$X
  mu0 <- beta0[m] + beta1[m] * 0 + beta2[m] * dat$X

  y1 <- rnorm(n, mu1, sigma[m])
  y0 <- rnorm(n, mu0, sigma[m])

  psi1_sim[m] <- mean(y1)
  psi0_sim[m] <- mean(y0)
}

delta_sim <- psi1_sim - psi0_sim
```

## Posterior predictive summary

```{r}
c(
  mean_effect = mean(delta_sim),
  lower       = quantile(delta_sim, 0.025),
  upper       = quantile(delta_sim, 0.975)
)
```

## What each step means

The code above has three layers.

### Causal identification layer

The g-formula says

$$
E(Y \mid do(T=t))
=
\int E(Y \mid T=t, X=x)\, f_X(x)\, dx.
$$

### Bayesian inference layer

The posterior distribution

$$
p(\theta \mid D)
$$

quantifies uncertainty about the outcome model parameters.

### Predictive layer

For each posterior draw, we construct hypothetical treated and untreated worlds and compute the average outcome in each world.

Thus:

- the **posterior distribution** handles uncertainty about unknown model parameters
- the **posterior predictive distribution** generates hypothetical outcomes
- **g-computation** averages those outcomes to produce causal estimates

## Why the duplicated-dataset intuition is correct

Your intuitive algorithm is:

1. keep the observed $X_i$
2. set treatment to 1 for everyone
3. simulate or predict outcomes
4. average them
5. set treatment to 0 for everyone
6. simulate or predict outcomes again
7. average them
8. compare the two averages

This is not merely similar to the g-formula. It **is** the g-formula in computational form.

The formal integral

$$
\int E(Y \mid T=t, X=x)\, f_X(x)\, dx
$$

is being approximated by the empirical average

$$
\frac{1}{n}\sum_{i=1}^n E(Y \mid T=t, X_i).
$$

So your intuitive predictive simulation story and the formal g-formula are the same method viewed from two angles.

## Extension to ETT

If the target is the effect of treatment on the treated, then we average over the covariate distribution of treated subjects only:

$$
ETT
=
\int
\left\{
E(Y \mid T=1, X=x) - E(Y \mid T=0, X=x)
\right\}
f_{X \mid T=1}(x)\, dx.
$$

In practice, that means repeating the same g-computation procedure but averaging over the treated subjects’ observed covariate values rather than over all subjects.

## Key take-home messages

- The **g-formula** is the causal averaging rule.
- The **posterior distribution** describes uncertainty about unknown model parameters.
- The **posterior predictive distribution** describes possible future outcomes under a treatment and covariate profile.
- Duplicating the dataset, fixing treatment, predicting outcomes, and averaging is exactly **Bayesian g-computation**.
- For continuous $X$, we do not use probabilities like $P(X=0)$ or $P(X=1)$; instead, we average over the empirical distribution of the observed covariates.

## Final summary

Bayesian causal inference via g-computation is conceptually simple. We fit a Bayesian outcome model to observational data, then construct hypothetical worlds in which every subject receives treatment 1 and, separately, treatment 0, while preserving each subject’s observed covariates. We predict or simulate outcomes in those two worlds, average outcomes across the target population, and compare the resulting intervention means. The g-formula provides the formal justification for this averaging step, while the posterior and posterior predictive distributions provide the Bayesian machinery for estimation and uncertainty propagation.




