---
title: "Lecture Notes on Power, Assurance, Conditional Power, Predictive Power, and Related Operating Characteristics"
author: "🌙☕📚💻"
date:  today
format: 
  html:
    theme: cyborg
    code-fold: true
    page-layout: article
    toc: true
    number-sections: true
    #toc-location: left
    self-contained-math: true
    embed-resources: true
    html-math-method: katex
execute: 
  echo: true
  eval: false 
  fig-height: 7
#bibliography: ../references.bib
#csl: https://www.zotero.org/styles/sage-vancouver-brackets
editor_options: 
  chunk_output_type: console
---

## Overview

These notes explain a family of operating characteristics used in clinical trial design and interim decision-making:

- classical power
- assurance
- conditional power
- predictive power or predictive probability of success
- conditional assurance
- Bayesian versions of power and assurance when success is defined through a posterior decision rule

A major source of confusion is that all of these quantities ask some version of the same question:

What is the probability that the study will eventually be declared successful?

However, they differ along two main dimensions:

1. whether the treatment effect parameter is fixed at a single value or integrated over uncertainty
2. whether the calculation is done before the trial starts or after interim data have been observed

A second source of confusion is that the final success criterion itself can be either:

- data-based or statistic-based, such as a final $Z$-test or a $p$-value threshold
- posterior-based, such as a posterior probability exceeding a threshold

These notes separate these issues carefully and explain both the mathematics and the simulation logic used in practice.

## General framework

Let

- $\theta$ denote the parameter of interest, for example a treatment effect
- $Y$ denote the full trial data
- $Y_1$ denote the data available at an interim analysis
- $Y_2$ denote the future data not yet observed, so that $Y = (Y_1, Y_2)$
- $D(Y)$ denote the final decision function, where
  $$
  D(Y) =
  \begin{cases}
  1, & \text{if the trial is declared successful} \\
  0, & \text{otherwise}
  \end{cases}
  $$

Thus every operating characteristic we discuss is ultimately some probability or expectation involving $D(Y)$.

## The two core dimensions

### Fixed effect versus integrating over effect uncertainty

At one extreme, we fix $\theta$ at a single value, such as a design alternative $\theta^\ast$.

At the other extreme, we acknowledge uncertainty about $\theta$ and average over a distribution:

- a prior distribution before the trial starts
- a posterior distribution after observing interim data

### Pre-trial versus interim

Before the trial starts, no data have been observed yet.

After an interim analysis, we condition on the observed interim data $Y_1 = y_1$ and ask about the remaining portion of the study.

These two dimensions generate the main family of operating characteristics discussed here.

## Trial success can be defined in two different ways

Before defining the operating characteristics, it is essential to distinguish two types of final success rules.

### Data-based or statistic-based success rule

A common frequentist-style rule is

$$
D(Y) = I\{ Z(Y) > c \},
$$

or equivalently

$$
D(Y) = I\{ p\text{-value}(Y) < \alpha \}.
$$

Here, once a full dataset $Y$ is available, we compute the final test statistic directly from the data and determine success immediately.

No posterior updating is required.

### Posterior-based success rule

A common Bayesian rule is

$$
D(Y) = I\{ P_A(\theta > \delta \mid Y) > \gamma \},
$$

where

- $\delta$ is a clinically meaningful threshold
- $\gamma$ is a posterior probability cutoff
- the subscript $A$ indicates that the posterior is computed using a prespecified analysis prior

In this case, once a full dataset $Y$ is available, we must first update the posterior and then determine success.

Thus, in practice:

- data-based rules use the hypothetical completed data directly
- posterior-based rules require the hypothetical completed data to be combined with an analysis prior to form a posterior

This distinction applies to every operating characteristic below.

## Continuous endpoint setup

To make the ideas concrete, we use a continuous endpoint setup throughout.

Suppose the parameter of interest is

$$
\theta = \mu_T - \mu_C,
$$

the difference in means between treatment and control.

Let the final estimator be $\hat{\theta}_N$, based on the total information or total sample size at the end of the study. Assume

$$
\hat{\theta}_N \mid \theta \sim N(\theta, V_N).
$$

Equivalently, define the final $Z$-statistic

$$
Z_N = \frac{\hat{\theta}_N}{\sqrt{V_N}}.
$$

For a one-sided superiority test, success under a data-based rule may be defined as

$$
D(Y) = I\{ Z_N > c \},
$$

where $c = z_{1-\alpha}$.

At an interim analysis, let the interim estimator be $\hat{\theta}_n$ with variance $V_n$, and let

$$
I_n = \frac{1}{V_n}, \qquad I_N = \frac{1}{V_N}.
$$

Then define

$$
Z_n = \hat{\theta}_n \sqrt{I_n}, \qquad Z_N = \hat{\theta}_N \sqrt{I_N}.
$$

A standard decomposition is

$$
Z_N
=
\sqrt{\frac{I_n}{I_N}} Z_n
+
\sqrt{\frac{I_N - I_n}{I_N}} Z_{\text{new}},
$$

where

$$
Z_{\text{new}} \mid \theta \sim N\left( \theta \sqrt{I_N - I_n}, 1 \right).
$$

This decomposition is central for conditional power and predictive power in the continuous normal case.

## Classical power

### Definition

Classical power fixes $\theta$ at a specific value and asks for the probability of final success.

Formally,

$$
Power(\theta) = P_\theta(D(Y) = 1).
$$

If the final rule is data-based with $D(Y) = I\{ Z_N > c \}$, then

$$
Power(\theta)
=
P_\theta\left( \frac{\hat{\theta}_N}{\sqrt{V_N}} > c \right).
$$

Since

$$
\frac{\hat{\theta}_N}{\sqrt{V_N}} \mid \theta
\sim
N\left( \frac{\theta}{\sqrt{V_N}}, 1 \right),
$$

we obtain

$$
Power(\theta)
=
1 - \Phi\left( c - \frac{\theta}{\sqrt{V_N}} \right).
$$

### Interpretation

Classical power answers:

If the true effect were exactly $\theta$, what is the probability that the final analysis would declare success?

This is a pre-trial quantity and does not condition on any observed interim data.

### Simulation in practice

If success is data-based:

1. fix $\theta = \theta^\ast$
2. simulate full trial datasets
   $$
   Y^{(m)} \sim p(Y \mid \theta^\ast), \qquad m = 1, \dots, M
   $$
3. compute the final test statistic from each dataset
4. define
   $$
   D^{(m)} = D\left( Y^{(m)} \right)
   $$
5. estimate power by
   $$
   \widehat{Power}(\theta^\ast)
   =
   \frac{1}{M} \sum_{m=1}^M D^{(m)}
   $$

If success is posterior-based:

1. fix $\theta = \theta^\ast$
2. simulate full trial datasets
3. for each simulated dataset, combine it with the analysis prior $\pi_A(\theta)$
4. compute the posterior
   $$
   \pi_A(\theta \mid Y^{(m)})
   $$
5. define success from the posterior rule
6. average the resulting indicators

Thus classical power does not inherently require or exclude posterior updating. It depends on the planned final decision rule.

## Assurance

### Definition

Assurance is the prior average of classical power.

If the true effect is assigned a design prior $\pi_D(\theta)$, then

$$
Assurance
=
\int Power(\theta) \, \pi_D(\theta) \, d\theta.
$$

Equivalently,

$$
Assurance
=
E_{\theta \sim \pi_D}\left[ Power(\theta) \right].
$$

### Interpretation

Assurance answers:

Before the trial starts, averaging over plausible true effects under the design prior, what is the probability that the study will eventually be declared successful?

Classical power conditions on a single effect value. Assurance does not.

### Why assurance is often lower than classical power

Suppose a trial is designed at a favorable effect size, such as $\theta^\ast = 0.40$, and the resulting classical power is high.

If the design prior places substantial mass on smaller effects, then averaging over that uncertainty may produce a much smaller assurance.

Thus assurance is often regarded as a more realistic pre-trial measure when the true effect is uncertain.

### Simulation in practice with a data-based final rule

1. draw a true effect from the design prior
   $$
   \theta^{(m)} \sim \pi_D(\theta)
   $$
2. simulate full trial data
   $$
   Y^{(m)} \sim p(Y \mid \theta^{(m)})
   $$
3. compute success directly from the full simulated data using the final planned statistic-based rule
4. define
   $$
   D^{(m)} = D\left( Y^{(m)} \right)
   $$
5. estimate assurance by
   $$
   \widehat{Assurance}
   =
   \frac{1}{M} \sum_{m=1}^M D^{(m)}
   $$

This is the version that is closest in spirit to frequentist trial evaluation, except that the design stage still averages over a prior distribution of $\theta$.

### Simulation in practice with a posterior-based final rule

In Bayesian design practice, one often separates the prior used for generating plausible true effects from the prior used in actual data analysis.

Let

- $\pi_D(\theta)$ be the design prior
- $\pi_A(\theta)$ be the analysis prior

Then the simulation algorithm is:

1. draw
   $$
   \theta^{(m)} \sim \pi_D(\theta)
   $$
2. simulate
   $$
   Y^{(m)} \sim p(Y \mid \theta^{(m)})
   $$
3. update using the analysis prior
   $$
   \pi_A(\theta \mid Y^{(m)}) \propto p(Y^{(m)} \mid \theta)\pi_A(\theta)
   $$
4. determine success using the posterior-based rule
5. average the success indicators

Thus there are two practical assurance pipelines:

- frequentist-like or data-based assurance
- fully Bayesian assurance using an analysis prior and posterior-based success rule

### Formal Bayesian assurance with separate design and analysis priors

If success is defined by

$$
D_A(Y) = I\{ P_A(\theta > \delta \mid Y) > \gamma \},
$$

then Bayesian assurance is

$$
Assurance_B
=
E_{\theta \sim \pi_D}
\left[
E_{Y \mid \theta}
\left[
D_A(Y)
\right]
\right].
$$

Equivalently,

$$
Assurance_B
=
\int
\int
D_A(Y)\, p(Y \mid \theta)\, \pi_D(\theta)\, dY \, d\theta.
$$

This is important in practice because the design prior and analysis prior do not have to be identical.

## Conditional power

### Definition

Conditional power is an interim quantity. It fixes the treatment effect at some value $\theta^\ast$ and asks for the probability of eventual success given the observed interim data.

Formally,

$$
CP(\theta^\ast; y_1)
=
P_{\theta^\ast}\left( D(Y_1, Y_2) = 1 \mid Y_1 = y_1 \right).
$$

### Interpretation

Conditional power answers:

Given the observed interim data, and assuming the true effect is exactly $\theta^\ast$, what is the probability that the final analysis will declare success?

### What is $\theta^\ast$?

This is simply a plug-in assumed value of the effect. Common choices include:

- the design alternative used in sample size planning
- the interim observed estimate
- a clinically meaningful effect size
- the posterior mean or posterior median, used merely as a plug-in number

What matters is that conditional power fixes the effect at a single value.

### Closed-form expression in the normal case

Using the normal setup and a data-based final rule $D(Y) = I\{ Z_N > c \}$, conditional power is

$$
CP(\theta^\ast; z_n)
=
P_{\theta^\ast}(Z_N > c \mid Z_n = z_n).
$$

Since

$$
Z_N
=
\sqrt{\frac{I_n}{I_N}} z_n
+
\sqrt{\frac{I_N - I_n}{I_N}} Z_{\text{new}},
$$

with

$$
Z_{\text{new}} \mid \theta^\ast
\sim
N\left( \theta^\ast \sqrt{I_N - I_n}, 1 \right),
$$

it follows that

$$
CP(\theta^\ast; z_n)
=
1 - \Phi\left(
\frac{
c
-
\sqrt{I_n/I_N}\, z_n
-
\theta^\ast (I_N - I_n)/\sqrt{I_N}
}{
\sqrt{(I_N - I_n)/I_N}
}
\right).
$$

### Simulation in practice

If success is data-based:

1. fix $\theta = \theta^\ast$
2. hold the observed interim data fixed at $Y_1 = y_1$
3. simulate future data
   $$
   Y_2^{(m)} \sim p(Y_2 \mid Y_1 = y_1, \theta^\ast)
   $$
4. create completed datasets
   $$
   Y^{(m)} = (y_1, Y_2^{(m)})
   $$
5. compute success directly from each completed dataset
6. average the success indicators

If success is posterior-based:

1. fix $\theta = \theta^\ast$
2. simulate future data
3. combine $(y_1, Y_2^{(m)})$ with the analysis prior
4. compute the hypothetical posterior
5. determine success from the posterior rule
6. average the indicators

Thus even conditional power may or may not include a posterior-updating step depending on the final success rule.

### Key point

Conditional power uses the conditional distribution of the future data given the observed interim data and a fixed effect:

$$
p(Y_2 \mid Y_1 = y_1, \theta^\ast).
$$

This is not the same as a posterior predictive distribution, because $\theta$ is not integrated out.

## Predictive power or predictive probability of success

### Definition

Predictive power averages conditional power over the posterior distribution of $\theta$ given the interim data.

Formally,

$$
PP(y_1)
=
P\left( D(Y_1, Y_2) = 1 \mid Y_1 = y_1 \right)
=
\int CP(\theta; y_1)\, p(\theta \mid y_1)\, d\theta.
$$

Equivalently,

$$
PP(y_1)
=
E_{\theta \mid y_1}\left[ CP(\theta; y_1) \right].
$$

### Interpretation

Predictive power answers:

Given the observed interim data, and accounting for current uncertainty about the true effect, what is the probability that the final analysis will declare success?

This is often called predictive probability of success.

### Formal nested representation

We can write

$$
PP(y_1)
=
\int
\int
D(y_1, y_2)\,
p(y_2 \mid y_1, \theta)\,
p(\theta \mid y_1)\,
dy_2\, d\theta.
$$

This shows the two layers clearly:

1. uncertainty in the remaining data $Y_2$
2. uncertainty in the true effect $\theta$

### Simulation in practice

If success is posterior-based:

1. compute the current posterior
   $$
   p(\theta \mid y_1)
   $$
2. for $m = 1, \dots, M$, draw
   $$
   \theta^{(m)} \sim p(\theta \mid y_1)
   $$
3. simulate future data
   $$
   Y_2^{(m)} \sim p(Y_2 \mid Y_1 = y_1, \theta^{(m)})
   $$
4. form completed datasets
   $$
   Y^{(m)} = (y_1, Y_2^{(m)})
   $$
5. combine $Y^{(m)}$ with the analysis prior
6. compute the hypothetical posterior
7. define success from the hypothetical posterior
8. estimate predictive power by averaging the success indicators

If success is data-based:

1. draw $\theta^{(m)}$ from the posterior
2. simulate $Y_2^{(m)}$
3. form $Y^{(m)} = (y_1, Y_2^{(m)})$
4. compute the final test statistic directly from $Y^{(m)}$
5. average the success indicators

Thus predictive power does not inherently require posterior updating at the final decision stage. It requires posterior averaging over $\theta$ at the interim stage. Whether an additional posterior update is needed for the success rule depends on how success is defined.

### Relationship to conditional power

The most important identity is

$$
PP(y_1)
=
\int CP(\theta; y_1)\, p(\theta \mid y_1)\, d\theta.
$$

Thus predictive power is the posterior average of conditional power.

## Conditional assurance

### Definition

Conditional assurance is an updated version of assurance after current evidence has been observed.

Let the current evidence be denoted by $Y_1 = y_1$, and let success refer to a future study or future portion of a development program. Then conditional assurance can be written as

$$
CA(y_1)
=
E_{\theta \mid y_1}
\left[
E_{Y_2 \mid y_1, \theta}
\left[
D(y_1, Y_2)
\right]
\right].
$$

Equivalently,

$$
CA(y_1)
=
\int CP(\theta; y_1)\, p(\theta \mid y_1)\, d\theta.
$$

### Why this looks identical to predictive power

Mathematically, conditional assurance and predictive power can take the same form.

The difference is often conceptual and contextual:

- predictive power usually refers to the chance that the current ongoing study will succeed at the final analysis
- conditional assurance often refers to the chance that a future study or future step in a development program will succeed, conditional on current evidence

Thus the event being predicted may differ even when the formula looks the same.

### Interpretation

Conditional assurance answers:

Given current evidence, averaging over the updated uncertainty about the true effect, what is the probability that the future planned study or future development step will succeed?

### Practical importance

Conditional assurance is particularly useful when the decision concerns whether to continue development, redesign a future study, or evaluate the value of a subsequent confirmatory trial.

## The core 2 by 2 structure

A useful way to organize the main quantities is the following.

| Time point | Fixed effect | Integrate over effect uncertainty |
|---|---|---|
| Pre-trial | Classical power | Assurance |
| Interim or current evidence | Conditional power | Predictive power or conditional assurance |

This table captures the most important logic.

## Direct data-based success versus posterior-based success

A major practical theme from these notes is that the same operating characteristic can be implemented under two different decision paradigms.

### Direct data-based success

Examples include:

- final $Z$-test significant
- final $p$-value below threshold
- final estimate above a clinically relevant margin

Pipeline:

$$
\text{hypothetical completed data}
\longrightarrow
\text{compute final statistic directly}
\longrightarrow
\text{success or failure}
$$

### Posterior-based success

Examples include:

- posterior probability $P(\theta > 0 \mid Y)$ exceeds $0.95$
- posterior probability $P(\theta > \delta \mid Y)$ exceeds a predefined threshold

Pipeline:

$$
\text{hypothetical completed data}
\longrightarrow
\text{combine with analysis prior}
\longrightarrow
\text{hypothetical posterior}
\longrightarrow
\text{success or failure}
$$

This distinction does not define whether a quantity is power, assurance, conditional power, or predictive power. It only determines how the success indicator $D(Y)$ is computed.

## Frequentist-like, hybrid, and fully Bayesian implementations

It is helpful to describe practical implementations using three labels.

### Frequentist-like

- success is defined directly from data or a final test statistic
- no analysis prior is needed in the final decision step

Example:

assurance computed by averaging over a design prior, but each hypothetical completed trial is judged using a final $Z$-test

This is not purely frequentist, because the design stage still averages over a prior distribution of $\theta$, but it is frequentist-like at the decision stage.

### Hybrid

- uncertainty in $\theta$ is handled through a prior or posterior in the design or predictive stage
- final success is still defined using a direct data-based statistic

Example:

predictive power where $\theta$ is drawn from the interim posterior, but final success is judged by whether the completed trial is statistically significant

### Fully Bayesian

- design or predictive averaging uses prior or posterior uncertainty about $\theta$
- final success is defined through a posterior-based criterion using an analysis prior

Example:

- Bayesian assurance with separate design prior and analysis prior
- predictive power under a posterior probability success rule

## The special issue of design prior and analysis prior

In Bayesian design practice, one must distinguish between:

- a design prior $\pi_D(\theta)$ used to represent plausible true effect values for simulation and operating characteristic evaluation
- an analysis prior $\pi_A(\theta)$ used in the actual analysis of each hypothetical completed dataset

These may be different.

### Why separate them?

The design prior is often intended to reflect realistic uncertainty about the true effect across possible worlds.

The analysis prior is often intended to represent the actual prior that will be used in the trial analysis and may be chosen for scientific, regulatory, or robustness reasons.

### Formal role of the two priors

For Bayesian assurance:

$$
\theta \sim \pi_D(\theta),
$$

then

$$
Y \mid \theta \sim p(Y \mid \theta),
$$

and finally

$$
\pi_A(\theta \mid Y) \propto p(Y \mid \theta)\pi_A(\theta).
$$

Success is then determined from the posterior under $\pi_A$.

This separation is important and should not be ignored in teaching Bayesian design.

## What if we simulate all of $Y_1$ and $Y_2$ after seeing interim data?

An idea discussed in our conversation was the following:

1. use the observed interim data $Y_1^{obs}$ to obtain a posterior for $\theta$
2. draw $\theta^{(m)}$ from that posterior
3. instead of keeping $Y_1^{obs}$ fixed, simulate a full hypothetical replicate dataset $(Y_1^{rep}, Y_2^{rep})$
4. evaluate success on that full hypothetical replicate dataset

This is a coherent procedure, but it is not standard predictive power for the ongoing trial.

### Why not?

Predictive power for the ongoing trial conditions on the actually observed interim data. Therefore, the observed interim data must remain fixed, and only the future unobserved data are simulated.

If one re-simulates $Y_1$, then one is no longer calculating

$$
P(\text{success in the ongoing trial} \mid Y_1 = y_1^{obs}),
$$

but rather something closer to a posterior-predictive probability of success for a new replicate full trial.

### Why the idea is still meaningful

If the model is adequate and the observed interim data are not unusual, then the posterior based on $Y_1^{obs}$ summarizes the interim evidence, and full replicate datasets generated from that posterior may resemble the observed trial in a distributional sense.

However, they do not condition on the realized interim dataset itself.

Thus this quantity answers a different question:

Given what the interim data have taught us about $\theta$, what is the chance that a new replicate full trial would succeed?

That is conceptually distinct from predictive power for the ongoing trial.

## What if we simulate only $Y_2$ and merge it with observed $Y_1$?

This is the standard practice for interim operating characteristics such as conditional power and predictive power.

The algorithm is:

1. hold $Y_1^{obs}$ fixed
2. simulate hypothetical future data $Y_2^{hyp}$
3. create
   $$
   Y^{full} = (Y_1^{obs}, Y_2^{hyp})
   $$
4. evaluate success on $Y^{full}$

If the final rule is data-based, success can be determined directly from the completed dataset.

If the final rule is posterior-based, then $Y^{full}$ must be combined with the analysis prior to create a hypothetical posterior before success is evaluated.

This is standard and appears throughout interim monitoring practice.

## Relationship identities worth remembering

The following identities are central.

### Assurance is the expectation of classical power under the design prior

$$
Assurance
=
E_{\theta \sim \pi_D}\left[ Power(\theta) \right].
$$

### Predictive power is the posterior expectation of conditional power

$$
PP(y_1)
=
E_{\theta \mid y_1}\left[ CP(\theta; y_1) \right].
$$

### Classical power is also the expectation of conditional power over interim realizations

If we fix $\theta$, then

$$
Power(\theta)
=
E_\theta\left[ CP(\theta; Y_1) \right].
$$

This identity is useful because it shows that when the interim data are allowed to vary again, the conditional quantity averages back to the unconditional one.

### Why classical power is the expectation of conditional power

A useful identity is

$$
Power(\theta) = E_\theta\left[ CP(\theta; Y_1) \right].
$$

This statement is correct, but it is often misunderstood because the expectation is **not** taken with respect to $\theta$. Here, the subscript $\theta$ means that the expectation is taken under the sampling distribution indexed by the fixed true effect $\theta$.

In other words, $\theta$ is fixed, while the interim data $Y_1$ are random.

To see this carefully, let $D(Y)$ be the indicator of final trial success:

$$
D(Y) =
\begin{cases}
1, & \text{if the final analysis declares success} \\
0, & \text{otherwise.}
\end{cases}
$$

Here, the full data are $Y = (Y_1, Y_2)$, where:

- $Y_1$ is the interim data
- $Y_2$ is the future data collected after the interim analysis

Under a fixed $\theta$, the classical power is

$$
Power(\theta) = P_\theta(D(Y)=1).
$$

Because $D(Y)$ is an indicator variable, this is the same as

$$
Power(\theta) = E_\theta[D(Y)].
$$

Now define the conditional power at interim value $Y_1 = y_1$ as

$$
CP(\theta; y_1) = P_\theta(D(Y)=1 \mid Y_1 = y_1).
$$

Again, since $D(Y)$ is an indicator, this can be written as

$$
CP(\theta; y_1) = E_\theta[D(Y) \mid Y_1 = y_1].
$$

If we now replace the realized interim value $y_1$ by the random interim variable $Y_1$, then

$$
CP(\theta; Y_1) = E_\theta[D(Y) \mid Y_1].
$$

Taking expectation on both sides gives

$$
E_\theta\left[ CP(\theta; Y_1) \right]
=
E_\theta\left[ E_\theta[D(Y) \mid Y_1] \right].
$$

By the law of total expectation,

$$
E_\theta\left[ E_\theta[D(Y) \mid Y_1] \right]
=
E_\theta[D(Y)].
$$

Therefore,

$$
Power(\theta) = E_\theta\left[ CP(\theta; Y_1) \right].
$$

This identity says that conditional power is a random quantity because it depends on the realized interim data. If we average it over all possible interim realizations under the fixed true effect $\theta$, we recover the ordinary unconditional classical power.

Equivalently, if $Y_1$ has density or mass function $p(y_1 \mid \theta)$, then

$$
Power(\theta)
=
\int CP(\theta; y_1)\, p(y_1 \mid \theta)\, dy_1,
$$

for continuous $Y_1$, or

$$
Power(\theta)
=
\sum_{y_1} CP(\theta; y_1)\, p(y_1 \mid \theta),
$$

for discrete $Y_1$.

It is important not to confuse this identity with the predictive power identity

$$
PP(y_1) = E_{\theta \mid y_1}\left[ CP(\theta; y_1) \right],
$$

because the two expectations are taken over different objects:

- in $Power(\theta) = E_\theta[CP(\theta; Y_1)]$, the expectation is over the sampling distribution of the interim data $Y_1$, with $\theta$ fixed
- in $PP(y_1) = E_{\theta \mid y_1}[CP(\theta; y_1)]$, the expectation is over the posterior distribution of $\theta$, with the interim data fixed at $y_1$

Thus the first is a frequentist averaging identity over possible interim datasets, whereas the second is a Bayesian averaging identity over posterior uncertainty in the treatment effect.


## A worked continuous normal example

Suppose the final estimator satisfies

$$
\hat{\theta}_N \mid \theta \sim N(\theta, 0.10^2),
$$

and the trial is declared successful if

$$
\hat{\theta}_N > 1.96 \times 0.10 = 0.196.
$$

### Classical power at a fixed effect

If $\theta^\ast = 0.30$, then

$$
Power(0.30)
=
P(\hat{\theta}_N > 0.196 \mid \theta = 0.30)
=
1 - \Phi\left( \frac{0.196 - 0.30}{0.10} \right).
$$

Thus

$$
Power(0.30)
=
1 - \Phi(-1.04)
=
\Phi(1.04)
\approx 0.851.
$$

### Assurance under a design prior

Suppose

$$
\theta \sim N(0.20, 0.15^2).
$$

Then under the prior predictive distribution,

$$
\hat{\theta}_N \sim N(0.20, 0.10^2 + 0.15^2).
$$

Therefore,

$$
Assurance
=
P(\hat{\theta}_N > 0.196)
=
1 - \Phi\left(
\frac{0.196 - 0.20}{\sqrt{0.10^2 + 0.15^2}}
\right).
$$

Since

$$
\sqrt{0.10^2 + 0.15^2}
=
\sqrt{0.0325}
\approx 0.1803,
$$

we obtain

$$
Assurance
\approx
1 - \Phi(-0.0222)
\approx
0.509.
$$

This illustrates how assurance can be much lower than classical power at a favorable fixed effect.

## Practical simulation pseudo-code

### Classical power with a data-based final rule

```{r}
M <- 10000
theta_star <- 0.30
success <- logical(M)

for (m in 1:M) {
  y_full <- simulate_full_trial(theta = theta_star)
  z_final <- final_test_statistic(y_full)
  success[m] <- (z_final > qnorm(0.975))
}

power_est <- mean(success)
power_est
```

### Assurance with a design prior and a data-based final rule

```{r}
M <- 10000
success <- logical(M)

for (m in 1:M) {
  theta_m <- rnorm(1, mean = 0.20, sd = 0.15)
  y_full <- simulate_full_trial(theta = theta_m)
  z_final <- final_test_statistic(y_full)
  success[m] <- (z_final > qnorm(0.975))
}

assurance_est <- mean(success)
assurance_est
```

### Conditional power with a posterior-based final rule

```{r}
M <- 10000
theta_star <- 0.25
success <- logical(M)

y1_obs <- observed_interim_data()

for (m in 1:M) {
  y2_hyp <- simulate_future_data(theta = theta_star, y1 = y1_obs)
  y_full <- merge_interim_and_future(y1_obs, y2_hyp)

  posterior <- update_posterior_analysis_prior(y_full)
  post_prob <- posterior_prob_theta_greater_than_delta(posterior, delta = 0)

  success[m] <- (post_prob > 0.95)
}

conditional_power_est <- mean(success)
conditional_power_est
```

### Predictive power with a posterior-based final rule

```{r}
M <- 10000
success <- logical(M)

y1_obs <- observed_interim_data()
posterior_interim <- update_posterior_analysis_prior(y1_obs)

for (m in 1:M) {
  theta_m <- draw_from_posterior(posterior_interim)
  y2_hyp <- simulate_future_data(theta = theta_m, y1 = y1_obs)
  y_full <- merge_interim_and_future(y1_obs, y2_hyp)

  hypothetical_posterior <- update_posterior_analysis_prior(y_full)
  post_prob <- posterior_prob_theta_greater_than_delta(hypothetical_posterior, delta = 0)

  success[m] <- (post_prob > 0.95)
}

predictive_power_est <- mean(success)
predictive_power_est
```

### Bayesian assurance with separate design and analysis priors

```{r}
M <- 10000
success <- logical(M)

for (m in 1:M) {
  theta_m <- draw_from_design_prior()
  y_full <- simulate_full_trial(theta = theta_m)

  posterior <- update_with_analysis_prior(y_full)
  post_prob <- posterior_prob_theta_greater_than_delta(posterior, delta = 0)

  success[m] <- (post_prob > 0.95)
}

bayesian_assurance_est <- mean(success)
bayesian_assurance_est
```

## Common misunderstandings

### Misunderstanding 1

Predictive power always requires a posterior update after simulating hypothetical future data.

This is false.

Predictive power requires posterior averaging over $\theta$ at the interim stage. But whether a second posterior update is needed after simulating future data depends on whether the final success rule is posterior-based.

### Misunderstanding 2

Assurance is always Bayesian and always requires an analysis prior.

This is false.

Assurance is Bayesian in the sense that it averages over uncertainty about the true effect, but the final success rule may be either:

- data-based, which does not require an analysis prior
- posterior-based, which does require an analysis prior

### Misunderstanding 3

Conditional assurance and predictive power are always different quantities.

This is false.

They can have the same mathematical form. The distinction is often contextual:

- predictive power is usually framed for the ongoing trial
- conditional assurance is often framed for a future study or future development step conditional on current evidence

### Misunderstanding 4

If one re-simulates both $Y_1$ and $Y_2$ after observing interim data, that is still predictive power for the current trial.

This is false.

That procedure produces something closer to a posterior-predictive success probability for a replicate full trial, not predictive power for the ongoing trial conditional on its actually observed interim data.

## Summary table

| Measure | Time point | Effect handling | Data simulated | Typical notation |
|---|---|---|---|---|
| Classical power | Pre-trial | Fix $\theta$ | Full trial data | $Power(\theta)$ |
| Assurance | Pre-trial | Integrate over design prior | Full trial data | $Assurance$ |
| Conditional power | Interim | Fix $\theta^\ast$ | Future data only | $CP(\theta^\ast; y_1)$ |
| Predictive power | Interim | Integrate over posterior | Future data only | $PP(y_1)$ |
| Conditional assurance | Current evidence or interim | Integrate over updated effect uncertainty | Future study or future data | $CA(y_1)$ |

## Final conceptual map

The cleanest way to remember all of this is:

### Pre-trial

- fix $\theta$
  $$
  \to \text{Classical power}
  $$
- integrate over prior uncertainty in $\theta$
  $$
  \to \text{Assurance}
  $$

### Interim or current evidence

- fix $\theta$
  $$
  \to \text{Conditional power}
  $$
- integrate over updated uncertainty in $\theta$
  $$
  \to \text{Predictive power or conditional assurance}
  $$

Then separately ask:

Is success defined directly from the completed data, or through a posterior computed using an analysis prior?

That second question determines the practical computation pipeline.

## Closing remark

These operating characteristics are best understood not as isolated formulas, but as a family generated by two design choices:

1. whether the effect is fixed or integrated over uncertainty
2. whether the calculation is pre-trial or conditional on current evidence

Once those two choices are clear, the remaining details follow systematically.




